<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Package Demo: raincpc}
-->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

# Introduction to *raincpc*

The Climate Prediction Center's ([CPC](www.cpc.ncep.noaa.gov)) global rainfall data, 1979 to present, 0.5 degree resolution (approximately 50 km), is one of the few high quality, long-term, observation-based, daily rainfall products available for free. Although raw data is available at [CPC's ftp site](ftp.cpc.ncep.noaa.gov/precip/CPC_UNI_PRCP/GAUGE_GLB/), obtaining and processing the data is not straightforward. 

Some issues with the raw CPC data:

* too many files (365/366 files per year * 36 years, separate folder for each year)
* each file has 360 rows and 720 columns
* file naming conventions have changed over time - one format prior to 2006 and several different formats afterwards
* file formats have changed over time - gzipped files prior to 2008 and plain binary files afterwards
* downloading multiple files simultaneously from the CPC ftp site, for instance using wget, does not seem to work properly
* there is no software readily available to easily process/visualize the data

The package *raincpc* addresses the above drawbacks by providing functionality to download, process and visualize the raw rainfall data from CPC. Following are some examples.

# Using *raincpc*

After installing the package, load the package along with *SDMTools*, *raster*, and *ggplot2*.

```{r, message = FALSE}
library(raincpc)
library(SDMTools)
library(raster)
library(ggplot2)
```

*raincpc* comes with the function `cpc_get_rawdata` to download the data from CPC. Following example extracts rainfall data for June 15-19 of 2014.

```{r}
cpc_get_rawdata(2014, 6, 15, 2014, 6, 19) 
```

Once the data is downloaded, the function `cpc_read_rawdata` can be used to read the binary data and convert it to a raster for analysis and graphics.

```{r}
rain1 <- cpc_read_rawdata(2014, 6, 15)
print(rain1)
```

Extract the rainfall for all the five days, and then compute the total rainfall for June 15-19. This rainfall total is used in the following graphics.

```{r}
rain2 <- cpc_read_rawdata(2014, 6, 16)
rain3 <- cpc_read_rawdata(2014, 6, 17)
rain4 <- cpc_read_rawdata(2014, 6, 18)
rain5 <- cpc_read_rawdata(2014, 6, 19)

rain_tot <- rain1 + rain2 + rain3 + rain4 + rain5
print(rain_tot)
```


# Graphics

Here is the plot of total rainfall for June 15-19 2014.

```{r}
png("gfx_rain1.png", width = 800, height = 600)
plot(rain_tot, 
     breaks = c(0, 80, 160, 240, 320),
     col = c("grey", "red", "green", "blue"), 
     main = "Rainfall (mm) June 15 - June 19 2014")
garbage <- dev.off()
```
![figure 1] [fig1]

Although it is easy to display the data using the plot method for a RasterLayer, the quality of the graphic is not great. Here is a plot of the same data using ggplot2. First, create a function to convert the data in a ggplot-friendly format.

```{r}
raster_ggplot <- function(rastx) {
  require(SDMTools)

  stopifnot(class(rastx) == "RasterLayer")
  
  gfx_data <- getXYcoords(rastx)
  # lats need to be flipped
  gfx_data <- expand.grid(lons = gfx_data$x, lats = rev(gfx_data$y), 
                            stringsAsFactors = FALSE, KEEP.OUT.ATTRS = FALSE)
  gfx_data$rain <- rastx@data@values
  
  return (gfx_data)
}
```

Use the above function to plot the RasterLayer object using ggplot2.

```{r}
rain_gg <- raster_ggplot(rain_tot)

rain_gg$rain_chunks <- cut(rain_gg$rain, breaks = c(0, 80, 160, 240, 320), 
                         include.lowest = TRUE)

gfx_gg <- ggplot(data = rain_gg)
gfx_gg <- gfx_gg + geom_raster(aes(lons, lats, fill = rain_chunks))
gfx_gg <- gfx_gg + scale_fill_manual(values = c("grey", "red", "green", "blue"))
gfx_gg <- gfx_gg + theme(axis.text = element_blank(), axis.ticks = element_blank())
gfx_gg <- gfx_gg + labs(x = NULL, y = NULL, fill = "Rain (mm)")
gfx_gg <- gfx_gg + ggtitle("Rainfall June 15 - June 19 2014")
  
png("gfx_rain2.png", width = 800, height = 600)
plot(gfx_gg)
garbage <- dev.off()
```
![figure 2] [fig2]


# Extracting Regional Rainfall

*SDMTools* has functionality to extract regional data from the RasterLayer object. In order to extract the data, say over Southeast Asia, specify a lat-lon bounding box as a data frame and extract the data for this box. 

```{r}
lon_vals <- seq(100.75, 130.75, 0.5)
lat_vals <- seq(-10.75, 20.75, 0.5)
reg_box <- expand.grid(lons = lon_vals, lats = lat_vals, 
                            stringsAsFactors = FALSE, KEEP.OUT.ATTRS = FALSE)
reg_box$rain <- extract.data(reg_box, rain_tot)
reg_box$rain_chunks <- cut(reg_box$rain, breaks = c(0, 80, 160, 240, 320), 
                         include.lowest = TRUE)
```

Use ggplot to plot the regional rainfall over Southeast Asia.

```{r}
gfx_gg <- ggplot(data = reg_box)
gfx_gg <- gfx_gg + geom_raster(aes(lons, lats, fill = rain_chunks))
gfx_gg <- gfx_gg + scale_fill_manual(values = c("grey", "red", "green", "blue"))
gfx_gg <- gfx_gg + theme(axis.text = element_blank(), axis.ticks = element_blank())
gfx_gg <- gfx_gg + labs(x = NULL, y = NULL, fill = "Rain (mm)")
gfx_gg <- gfx_gg + ggtitle("Rainfall over Southeast Asia June 15 - June 19 2014")
  
png("gfx_rain3.png")
plot(gfx_gg)
garbage <- dev.off()
```
![figure 3] [fig3]

[fig1]: gfx_rain1.png "figure 1" 
[fig2]: gfx_rain2.png "figure 2" 
[fig3]: gfx_rain3.png "figure 3" 
